Data preprocessing: raw data to processed data

### 2023-04-25 manuscript/manuscript.Rmd

keywords          : "Blind data analysis; preregistration; ALSPAC; meta-research; open science; Explore and Confirm Analysis Workflow (ECAW)"
wordcount         : "3546"
bibliography      : "ecaw-ref.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
appendix: "supplementary_materials.Rmd"

>>>>

keywords          : "Blind data analysis; preregistration; ALSPAC; meta-research; open science; Explore and Confirm Analysis Workflow (ECAW)"
wordcount         : "4020"
bibliography      : "ecaw-ref.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
appendix: "supplementary_materials.Rmd"

### 2023-04-25 manuscript/manuscript.tex

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Blind data analysis; preregistration; ALSPAC; meta-research; open science; Explore and Confirm Analysis Workflow (ECAW)\newline\indent Word count: 3546}
\usepackage{lineno}

@@

Of 1148 emails sent, 1094 went through and 54 bounced. The survey was completed 107 times and partially completed 21 times, leading to a response rate of 10\% for complete surveys and 2\% for incomplete surveys.\footnote[1]{The ALSPAC mailing list has been maintained as a record of collaborators for many years and is constantly added to. The ALSPAC team only removes email addresses from their mailing list if someone explicitly requests this action. Thus, their list likely includes several email addresses that are no longer monitored. For example, we received one email reply stating that the recipient hasn’t been active in research for 30 years. It is also possible that some researchers have multiple email addresses on the mailing list (e.g., because they moved institutions). These two factors may have deflated the response rate.} The median time taken for complete survey responses was 7.20 minutes (IQR: 4.30 to 13.10). This manuscript presents the results for complete surveys. Supplementary Material D presents the results with partially complete surveys included.

@@

Respondents had published a median of NA (IQR 2 to 25) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: R (n = 65), Stata (n = 48), SPSS (n = 17), SAS (n = 15), Python (n = 6), Mplus (n = 3), Bash (n = 2), MATLAB (n = 1), Nextflow (n = 1), and plink2 (n = 1) (Supplementary Table C1)\footnote[2]{Participants could select multiple responses to this survey question.}. 61\% (62/101) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; 6\% (6/101) reported being less concerned (Supplementary Figure C2).

@@

Most respondents agreed that studies that analyze preexisting observational datasets are trustworthy (72\%; 77/107) and reproducible (79\%; 84/107) (Figure 2, top panel). At the same time, many agreed that a study using an ECAW would be \emph{more} trustworthy (70\%; 71/101) and \emph{more} reproducible (69\%; 70/102) compared to a typical study using preexisting observational data (Figure 1, bottom panel).

@@

\caption{\textbf{Figure 1. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.} The survey defined trustworthy as ``meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust''. The survey defined reproducible ``in the sense that other researchers re-analysing the data with the same research question would produce similar results.'' For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option \emph{``I don't understand the question''}. The bar charts in the bottom panel excluded missing responses (n = 2; 2) and responses of \emph{``I don't understand the question''} (n = 4; 3).}\label{fig:typicallyEcawPlot}
\end{figure}

Over half of respondents reported that their studies using preexisting observational data are preregistered never or almost never (35\%; 37/107), or sometimes (24\%; 26/107) (Figure 2). About half reported sharing their analysis scripts never or almost never (20\%; 21/107), or sometimes (31\%; 33/107). 75\% (80/107) reported that they never or almost never blind the data analyst. Almost all respondents answered that they use both exploratory (91\%; 98/107) and confirmatory (85\%; 92/107) analyses at least sometimes.

@@

\caption{\textbf{Figure 3. Responses to survey questions about using ECAWs.} These bar charts exclude missing values (\emph{n} = 3; 3; 3; 3; respectively from top to bottom), responses of \emph{``I don't understand the question''} (\emph{n} = 0; 4; 1; 1), and responses of \emph{``Unsure''} (\emph{n} = 2; 5; 6; 8). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word ``less'' in the first question \protect\footnotemark[3]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term ``study'' \protect\footnotemark[4].}\label{fig:alspacPlot}
\end{figure}

@@

\caption{\label{tab:languageTable}\textbf{Table C1. ``What programming language or software do you use for your analyses of preexisting observational data? (you may select multiple answers)''}}
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrr}
\toprule
\textbf{Programming language} & \textbf{N} & \textbf{Percentage of respondents}\\
\midrule
R & 65 & 61\\
\cmidrule{1-3}
Stata & 48 & 45\\
\cmidrule{1-3}
SPSS & 17 & 16\\
\cmidrule{1-3}
SAS & 15 & 14\\
\cmidrule{1-3}
Python & 6 & 6\\
\cmidrule{1-3}
Missing & 4 & 4\\
\cmidrule{1-3}
Mplus & 3 & 3\\
\cmidrule{1-3}
Bash & 2 & 2\\
\cmidrule{1-3}
MATLAB & 1 & 1\\
\cmidrule{1-3}
Nextflow & 1 & 1\\
\cmidrule{1-3}
plink2 & 1 & 1\\
\bottomrule
\end{tabular}}
\end{table}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Blind data analysis; preregistration; ALSPAC; meta-research; open science; Explore and Confirm Analysis Workflow (ECAW)\newline\indent Word count: 4020}
\usepackage{lineno}

@@

Of 1148 emails sent, 1094 went through and 54 bounced. The survey was completed 103 times and partially completed 17 times, leading to a response rate of 9\% for complete surveys and 2\% for incomplete surveys.\footnote[1]{The ALSPAC mailing list has been maintained as a record of collaborators for many years and is constantly added to. The ALSPAC team only removes email addresses from their mailing list if someone explicitly requests this action. Thus, their list likely includes several email addresses that are no longer monitored. For example, we received one email reply stating that the recipient hasn’t been active in research for 30 years. It is also possible that some researchers have multiple email addresses on the mailing list (e.g., because they moved institutions). These two factors may have deflated the response rate.} The median time taken for complete survey responses was 7.40 minutes (IQR: 4.60 to 13.10). This manuscript presents the results for complete surveys. Supplementary Material D presents the results with partially complete surveys included.

@@

Respondents had published a median of NA (IQR 2 to 26.20) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: R (n = 65), Stata (n = 48), SPSS (n = 17), SAS (n = 15), Python (n = 6), Mplus (n = 3), Bash (n = 2), MATLAB (n = 1), Nextflow (n = 1), and plink2 (n = 1) (Supplementary Table C1)\footnote[2]{Participants could select multiple responses to this survey question.}. 62\% (62/100) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; 6\% (6/100) reported being less concerned (Supplementary Figure C2).

@@

Most respondents agreed that studies that analyze preexisting observational datasets are trustworthy (72\%; 74/103) and reproducible (79\%; 81/103) (Figure 2, top panel). At the same time, many agreed that a study using an ECAW would be \emph{more} trustworthy (70\%; 70/100) and \emph{more} reproducible (68\%; 69/101) compared to a typical study using preexisting observational data (Figure 1, bottom panel).

@@

\caption{\textbf{Figure 1. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.} The survey defined trustworthy as ``meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust''. The survey defined reproducible ``in the sense that other researchers re-analysing the data with the same research question would produce similar results.'' For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option \emph{``I don't understand the question''}. The bar charts in the bottom panel excluded missing responses (n = ; ) and responses of \emph{``I don't understand the question''} (n = 3; 2).}\label{fig:typicallyEcawPlot}
\end{figure}

Over half of respondents reported that their studies using preexisting observational data are preregistered never or almost never (36\%; 37/103), or sometimes (25\%; 26/103) (Figure 2). About half reported sharing their analysis scripts never or almost never (20\%; 21/103), or sometimes (32\%; 33/103). 77\% (79/103) reported that they never or almost never blind the data analyst. Almost all respondents answered that they use both exploratory (93\%; 96/103) and confirmatory (87\%; 90/103) analyses at least sometimes.

@@

\caption{\textbf{Figure 3. Responses to survey questions about using ECAWs.} These bar charts exclude missing values (\emph{n} = 0; 0; 0; 0; respectively from top to bottom), responses of \emph{``I don't understand the question''} (\emph{n} = 0; 4; 1; 1), and responses of \emph{``Unsure''} (\emph{n} = 2; 5; 6; 8). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word ``less'' in the first question \protect\footnotemark[3]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term ``study'' \protect\footnotemark[4].}\label{fig:alspacPlot}
\end{figure}

@@

\caption{\label{tab:languageTable}\textbf{Table C1. ``What programming language or software do you use for your analyses of preexisting observational data? (you may select multiple answers)''}}
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrr}
\toprule
\textbf{Programming language} & \textbf{N} & \textbf{Percentage of respondents}\\
\midrule
R & 65 & 63\\
\cmidrule{1-3}
Stata & 48 & 47\\
\cmidrule{1-3}
SPSS & 17 & 17\\
\cmidrule{1-3}
SAS & 15 & 15\\
\cmidrule{1-3}
Python & 6 & 6\\
\cmidrule{1-3}
Mplus & 3 & 3\\
\cmidrule{1-3}
Bash & 2 & 2\\
\cmidrule{1-3}
MATLAB & 1 & 1\\
\cmidrule{1-3}
Nextflow & 1 & 1\\
\cmidrule{1-3}
plink2 & 1 & 1\\
\bottomrule
\end{tabular}}
\end{table}

### 2023-04-25 preprocessing/ecaw_raw_processed_preprocessing.Rmd

@@

We only include full responses in the final analysis as preregistered (see the full text of the preregistration [https://osf.io/g2fw5](https://osf.io/g2fw5)). There were `r nrow(dplyr::filter(raw, !finished))` respondents who did not finish the survey.

```{r}
raw <- 
  raw %>% 
  dplyr::filter(finished)
```

These responses will be included in an exploratory analysis that will be published in supplementary materials D.

At the end of the exclusion we had `r nrow(raw)` responses remaining for the analysis.

```{r}
raw <-
  raw %>% 
    # dropping those questions that were open ended or conditional on other responses or irrelevant
    # flag those responses that has any missing values
    mutate(
      has_missing = case_when(
        if_any(
          c(
            -start_date,
            -end_date,
            -status,
            -progress,
            -duration_in_seconds,
            -finished,
            -recorded_date,
            -distribution_channel,
            -user_language,
            -programming_language_other,
            -unsure_explain,
            -drawback_ecaw,
            -suggestions,
            -comments
            ),
          .fns = is.na) ~ TRUE,
        TRUE ~ FALSE
        )
      )
```

There were `r filter(raw, has_missing) %>% nrow()` respondents who had at least one missing value in the main questions of interests.

@@

On median it took `r median(raw$duration_in_seconds) / 60` to finish the survey to the participants in hour sample.

@@

The survey contained 4 open-ended questions (*unsure_explain*, *drawback_ecaw*, *suggestions*, *comments*) that we analyzed qualitatively as preregistered (see page X line Y of the preregistration).

@@

For creating the codebook we will use the codebook created for the source data.

@@

# Save the codebook
readr::write_csv(codebook, here("data/raw/ecaw_raw_codebook.csv"))
```

>>>>

```{r}
raw <-
  raw %>% 
    # dropping those questions that were open ended or conditional on other responses or irrelevant
    # flag respondents with at least one missing values from a closed-ended question
    mutate(
      has_missing = case_when(
        if_any(
          c(
            -start_date,
            -end_date,
            -status,
            -progress,
            -duration_in_seconds,
            -finished,
            -recorded_date,
            -distribution_channel,
            -user_language,
            -programming_language_other,
            -unsure_explain,
            -drawback_ecaw,
            -suggestions,
            -comments
            ),
          .fns = is.na) ~ TRUE,
        TRUE ~ FALSE
        )
      )
```

@@

We only include data from participants who completed the survey in the final analysis, as preregistered (see the full text of the preregistration [https://osf.io/g2fw5](https://osf.io/g2fw5)). There were `r nrow(dplyr::filter(raw, !finished))` respondents who did not complete the survey.

However, these responses will be included in a supplementary analysis that will be reproted in Supplementary Material D. So we save them separately before exclusion.

```{r}
processed_all <- 
  raw %>%
  mutate(duration_in_mins = duration_in_seconds / 60) %>%
  select(
    -start_date,
    -end_date,
    -status,
    -progress,
    -finished,
    -duration_in_seconds,
    -recorded_date,
    -distribution_channel,
    -user_language
  ) %>% 
  # remove participants who only opened the survey but did not answer any questions. All participants who had NA for the first question (typically_trustworthy), had NAs for all questions
  filter(!is.na(typically_trustworthy))

processed_all <- 
  processed_all

readr::write_csv(processed_all, here("data/processed/ecaw_processed_all_data.csv"))
```

The full dataset that we are going to use to calculate the results in Supplement D includes `r nrow(processed_all)` responses.

We create a simple codebook for this dataset.

```{r}
processed_all_codebook <-
  read_csv(here("data/source/ecaw_221102_source_codebook.csv")) %>% 
  select(-original_var_name) %>% 
  rename(var_name = new_var_name) %>% 
  filter(var_name %in% colnames(processed_all))

# Adding new column descriptions
new_row1 <- tibble(
  description = "True if respondent has any missing responses.",
  var_name = "has_missing"
)
new_row2 <- tibble(
  description = "Time spent from opening the survey to closing it in minutes.",
  var_name = "duration_in_mins"
)

codebook <- bind_rows(processed_all_codebook, new_row1, new_row2)

readr::write_csv(processed_all_codebook, here("data/processed/ecaw_processed_all_codebook.csv"))
```

Excluding the respondents who did not finish the questionnaire.

```{r}
raw <- 
  raw %>% 
  dplyr::filter(finished) %>%
  filter(!(response_id %in% c("R_2ZVO2mDq0eBPYDn",
                              "R_3qr9uw1HGF53UiJ",
                              "R_3iQHcAwCVxBXAEd",
                              "R_1HqrkliX8SXIZqS")
  )
  )
# We did a brief visual inspection of the data and feel these first three respondents fit better as partially completed surveys. Although they clicked through to the end of the survey, they did no answer questions after a certain point. For example, one respondent only answered the first two quetsions and clicked through to the end of the survey in 49 seconds.
# The fourth response appears to be non-sincere. All their responses were neutral, several were skipped, and it was completed in 146 seconds. This respondent may have simply wanted to see the survey. 
```

At the end of the exclusion we had `r nrow(raw)` responses remaining for the analysis. There were `r filter(raw, has_missing) %>% nrow()` respondents who had at least one missing value for a closed-ended question.

@@

The median time to complete the survey was `r median(raw$duration_in_seconds) / 60`.

@@

The survey contained 4 open-ended questions (*unsure_explain*, *drawback_ecaw*, *suggestions*, *comments*) that we analyzed qualitatively, as preregistered.

@@

For creating the codebook for the processed data we will use the codebook created for the source data.

@@

# Adding new column descriptions
new_row1 <- tibble(
  description = "True if respondent has any missing responses.",
  var_name = "has_missing"
)
new_row2 <- tibble(
  description = "Time spent from opening the survey to closing it in minutes.",
  var_name = "duration_in_mins"
)

codebook <- bind_rows(codebook, new_row1, new_row2)

# Save the codebook
readr::write_csv(codebook, here("data/processed/ecaw_processed_codebook.csv"))
```
